---
title: "Class 12"
author: "Abigail Perez"
format: pdf
editor: visual
---

## 1. Background

For this class session we will:

Open a new RStudio Project and Quarto document for today’s class; Review how to install both Bioconductor and CRAN packages; Explore the Himes et al. gene expression data using base R, dplyr and ggplot2 package functions; Perform a detailed differential gene expression analysis with the DESeq2 package. Render a reproducible PDF report of your work with answers to all questions below.

## 2. Bioconductor setup

```{r}
library(BiocManager)
library(DESeq2)
library(ggplot2)
library(dplyr)
```

## 3. Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")

head(counts)
head(metadata)
```

> **Q1. How many genes are in this dataset?**

> **Answer:** There are 38,694 genes in this dataset, as determined by nrow(counts).

```{r}
nrow(counts)
```

> **Q2. How many ‘control’ cell lines do we have?**

> **Answer:** There are 4 control cell lines.

```{r}
sum(metadata$dex == "control")
```

## 4. Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

> **Q3. How would you make the above code in either approach more robust? Is there a function that could help here?**

> **Answer:** We can use rowSums(). This makes the code more generalizable if the number of control samples ever changes.

```{r}
control.mean <- rowSums(control.counts) / ncol(control.counts)
```

> **Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)**

> **Answer:**

```{r}
treated <- metadata[metadata[,"dex"] == "treated", ]
treated.counts <- counts[, treated$id]
treated.mean <- rowMeans(treated.counts)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

> **Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.**

> **Answer:**

```{r}
plot(meancounts[,1], meancounts[,2],
     xlab = "Control Mean", ylab = "Treated Mean")
```

> **Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom\_?() function would you use for this plot?**

> **Answer:** geom_point() — since this is a scatter plot comparing two continuous variables (control mean vs treated mean), each gene is represented as a single point.

```{r}
ggplot(meancounts, aes(x = control.mean, y = treated.mean)) + geom_point(alpha = 0.3) +
xlab("Control") + ylab("Treated")
```

> **Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?**

> **Answer:** The log = "xy" argument tells plot() to apply a logarithmic scale to both the x and y axes simultaneously. This is useful here because most genes cluster near zero, making the raw scale hard to read.

```{r}
# Base R:
plot(meancounts[,1], meancounts[,2], log = "xy",
     xlab = "log Control Mean", ylab = "log Treated Mean")

# ggplot:
# add: scale_x_continuous(trans="log2") + scale_y_continuous(trans="log2")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"] / meancounts[,"control.mean"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

```{r}
head(mycounts$log2fc)
nrow(mycounts)
```

> **Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?**

> **Answer:** Setting arr.ind = TRUE makes which() return row and column indices as a matrix rather than a single positional number, which is necessary since we are searching across two columns. We take the first column to extract the row indices corresponding to genes with zeros, then call unique() to avoid removing the same gene twice if it had zeros in both columns.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> **Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level?**

> **Answer:** There are 250 upregulated genes at the greater than 2 log2 fold change threshold, as determined by sum(up.ind).

```{r}
sum(up.ind)
```

> **Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?**

> **Answer:** There are 367 downregulated genes at the less than -2 log2 fold change threshold, as determined by sum(down.ind).

```{r}
sum(down.ind)
```

> **Q10. Do you trust these results? Why or why not?**

> **Answer:** No, because this naive approach compares group means without accounting for statistical significance, biological variability, or differences in sequencing depth, meaning many flagged genes could be false positives. DESeq2 is a much more reliable method as it normalizes for library size, models count dispersion, and applies proper statistical testing with adjusted p-values.

## 5. Setting up for DESeq

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

## 6. Principal Component Analysis (PCA)

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

```{r}
# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

## 7. DESeq analysis

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

## 8. Adding annotation data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
head(res)
```

> **Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res**$entrez, res$uniprot and res\$genename.

> **Answer:**

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

## 9. Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
BiocManager::install("EnhancedVolcano")
```

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

## 10. Pathway analysis

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

foldchanges <- res$log2FoldChange
names(foldchanges) <- res$entrez

keggres <- gage(foldchanges, gsets = kegg.sets.hs)
head(keggres$less, 3)

pathview(gene.data = foldchanges, pathway.id = "hsa05310")
```

> **Q12. Can you do the same procedure as above to plot the pathview figures for the top 2 down-reguled pathways?**

> **Answer:** Yes. From head(keggres\$less, 3), the top two downregulated pathways are Graft-versus-host disease (hsa05332) and Type I diabetes mellitus (hsa04940). We can visualize these using the same pathview() call with the corresponding pathway IDs, which downloads the KEGG pathway diagram and overlays our fold change data to show which genes in each pathway are up or downregulated in response to dexamethasone treatment.

```{r}
pathview(gene.data = foldchanges, pathway.id = "hsa05332")
pathview(gene.data = foldchanges, pathway.id = "hsa04940")
```

## OPTIONAL: Plotting counts for genes of interest

```{r}
i <- grep("CRISPLD2", res$symbol)
res[i,]
rownames(res[i,])

plotCounts(dds, gene = "ENSG00000103196", intgroup = "dex")

d <- plotCounts(dds, gene = "ENSG00000103196", intgroup = "dex", returnData = TRUE)

ggplot(d, aes(dex, count, fill = dex)) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("CRISPLD2")
```

```{r}
sessionInfo()
```
